### 2021-04-24 submission/manuscript.Rmd
Corrections after trying to render manuscript

---
title: ''
csl: mbio.csl #Get themes at https://github.com/citation-style-language/styles
fontsize: 11pt
output:
  word_document: default
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
  html_document:
    df_print: paged
geometry: margin=1.0in
bibliography: references.bib
---

@@

``` {r figure_5_permanova}
fig_5_permanova_results = read_tsv("data/process/post_CDI_PEG_permanova_stools.tsv")
r_sq_ordered_perm = permanova_results %>% arrange(desc(R2)) %>%
  select(`effects`, `R2`, `Pr(>F)`) %>%
  filter(`Pr(>F)` < .05) #filter for sig
fig_5_permanova_sig_P_val <- r_sq_ordered_perm[,3] %>% sum()
fig_5_permanova_sig_R2 <- r_sq_ordered_perm[,2] %>% sum() #0.9280951
```

\includegraphics{figure_5.pdf}
**Figure 5. 1-day PEG treatment post *C. difficile* challenge prolongs colonization regardless of whether an FMT is also administered.** All mice were administered 10 mg/kg clindamycin intraperitoneally (IP) 1 day before challenge with C. difﬁcile 630 spores on day 0. 15% PEG 3350 was administered in the drinking water for 1 day. Mice feces were collected daily through the end of the experiment (30 days post-infection for Clind. + 1-day PEG group, 15 days post-infection for all others). CFU quantification data was not available for each mouse due to stool sampling difficulties (particularly the day the mice came off of the PEG treatment) or early deaths. B. CFU/g of C. difficile stool measured over time via serial dilutions. C. Median and individual Shannon Diversity measured over time. D. Median and individual richness measured over time. B-D. Asterisks indicate time points with significant differences between groups using Kruskall-Wallis tests with a Benjamini-Hochberg correction for testing multiple times points. Black line represents the limit of detection for the first serial dilution. Background shading indicates PEG treatment period for applicable groups. Opacity of points reflects time point. The data presented are from a total of *** separate experiments.
\newpage

>>>>

---
title: ''
csl: mbio.csl #Get themes at https://github.com/citation-style-language/styles
fontsize: 11pt
output:
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
  word_document: default
  html_document:
    df_print: paged
geometry: margin=1.0in
bibliography: references.bib
---

@@

``` {r figure_5_permanova}
fig_5_permanova_results = read_tsv(here("data/process/post_CDI_PEG_permanova_stools.tsv"))
r_sq_ordered_perm = fig_5_permanova_results %>% arrange(desc(R2)) %>%
  select(`effects`, `R2`, `Pr(>F)`) %>%
  filter(`Pr(>F)` < .05) #filter for sig
fig_5_permanova_sig_P_val <- r_sq_ordered_perm[,3] %>% sum()
fig_5_permanova_sig_R2 <- r_sq_ordered_perm[,2] %>% sum() #0.9280951
```

\includegraphics{figure_5.pdf}
**Figure 5. 1-day PEG treatment post *C. difficile* challenge prolongs colonization regardless of whether an FMT is also administered.** All mice were administered 10 mg/kg clindamycin intraperitoneally (IP) 1 day before challenge with *C. difficile* 630 spores on day 0. 15% PEG 3350 was administered in the drinking water for 1 day. Mice feces were collected daily through the end of the experiment (30 days post-infection for Clind. + 1-day PEG group, 15 days post-infection for all others). CFU quantification data was not available for each mouse due to stool sampling difficulties (particularly the day the mice came off of the PEG treatment) or early deaths. B. CFU/g of *C. difficile* stool measured over time via serial dilutions. C. Median and individual Shannon Diversity measured over time. D. Median and individual richness measured over time. B-D. Asterisks indicate time points with significant differences between groups using Kruskall-Wallis tests with a Benjamini-Hochberg correction for testing multiple times points. Black line represents the limit of detection for the first serial dilution. Background shading indicates PEG treatment period for applicable groups. Opacity of points reflects time point. The data presented are from a total of *** separate experiments.
\newpage

### 2021-04-24 submission/manuscript.tex
Corrections after trying to render manuscript

\textbf{5-day laxative treatment leads to prolonged \emph{C. difficile}
colonization in mice.} We compared mice treated with the osmotic
laxative PEG 3350 to our standard 10 mg/kg clindamycin treatment, which
temporarily renders the mice susceptible to \emph{C. difficile}, with
mice typically clearing \emph{C. difficile} within 10 days
post-infection (9, 19). All PEG-treated mice were administered a 15\%
PEG solution in the drinking water for 5-days, one group was also
treated with clindamycin, and one group was allowed to recover for 10
days prior to challenge (Fig. 1A). PEG treatment resulted in weight loss
in all 3 groups of PEG-treated mice, with the greatest change in weight
observed on the fifth day of PEG treatment and the mice regained most of
the weight five days after treatment (Fig. 1B). After either PEG,
clindamycin, or PEG and clindamycin treatment all mice were challenged
with 10\textsuperscript{3} \emph{C. difficile} 630 spores. All
treatments rendered mice susceptible to \emph{C. difficile} colonization
(Fig. 1C), however PEG-treated mice remained colonized at a high level
through thirty days post-infection. In contrast, the clindamycin-treated
mice that cleared \emph{C. difficile} within ten days post-infection.
Surprisingly, PEG-treated mice were still susceptible to \emph{C.
difficile} infection after 10-days of recovery from treatment although
\emph{C. difficile} was not detectable in most of the group in the
initial five days post-infection (Fig. 1C). From 9 days post-infection
onward, the median \emph{C. difficile} stabilized for the 5-day PEG plus
10-day recovery group of mice and remained high through 30 days
post-infection (Fig. 1C). Thus, osmotic laxative treatment alone was
sufficient to render mice susceptible to prolonged \emph{C. difficile}
colonization and PEG-treated mice remained susceptible for up to ten
days post-treatment.

\textbf{5-day laxative treatment differentially disrupts the fecal
microbiota compared to clindamycin treatment.} Since laxatives and
clindamycin have previously been shown to disrupt the murine microbiota
(ref), we hypothesized the different \emph{C. difficile} colonization
dynamics between mice treated with laxatives versus clindamycin were due
to the two drugs having differential effects on the microbiota. We
profiled the stool microbiota over time by sequencing the V4 region of
the 16S rRNA gene to compare changes across treatment groups. We found
time and treatment group explained half of the observed variation
between fecal communities with most of the remaining variation explained
by interactions between treatment group and other experimental variables
including time, cage effects, and sequencing preparation (PERMANOVA
combined R\textsuperscript{2} = 0.95, \emph{P} \textless{} 0.001, Fig.
2A, Data Set S1, Sheet X). Cage effects refer to the well-documented
phenomenon that mice housed in the same cages have similar microbial
communities due to copraphagy, we tried to minimize the impact of cage
effects on our experiment by breaking up cagemates when assigning mice
to treatment groups and primarily housing only two mice per cage.
Importantly, although we conducted a total of 5 separate experiments,
the experiment number and it's interaction with treatment group was not
one of the variables that significantly explained the observed variation
in fecal communities (Data Set S1, Sheet X). Interestingly, none of the
treatment groups recovered to their baseline community structure either
10 or 30 days post-infection suggesting other community features besides
recovery to baseline were responsible for the prolonged \emph{C.
difficile} colonization in PEG-treated mice (Fig. 2B).

Next, we examined alpha diversity by looking at Shannon diversity index
over time, although both clindamycin and PEG treatments decreased
diversity, the diversity index was lower in the groups of mice that
received PEG treatment compared to those that received clindamycin
through thirty days post-infection (Fig. 2C). We next examined the
bacterial genera that shifted after PEG treatment by comparing the
baseline samples of mice treated with only PEG to samples from the same
mice one day post-treatment. We found 18 OTUs that were altered by PEG,
the majority of these OTUs decreased after PEG, but
\emph{Enterobacteriaceae} and \emph{Bacteroides} increased and the
increase in \emph{Bacteroides} was unique to PEG treatment as
Bacteroides actually decreased in clindamycin treated mice (Fig. 2D,
Data Set S1, Sheet X). Finally, we examined the bacteria that differ
across treatment groups over multiple timepoints. We found 24 were
different over multiple timepoints out of the 33 that were different
between treatment groups (Fig. 2E, Data Set S1, Sheet X). Thus, PEG has
a significant impact on the fecal microbiota that was maintained over
time and distinct from clindamycin treatment.

Surprisingly, \emph{C. difficile} was not immediately detectable in the
stools of the PEG-treated mice that were allowed to recover for 10 days
prior to challenge. We decided to examine the bacteria that changed
during the post-infection period when the group median \emph{C.
difficile} shifted from undetectable at 1 day post-infection to
detectable in the stool samples with the median stabilizing around 8
dayspost-infection (Fig. S1A). Interestingly, we found
\emph{Erysipelotrichaceae}, \emph{Enterobacteriaceae}, and
\emph{Akkermansia} were changing during the time period when \emph{C.
difficile} was becoming detectable in the stools (Fig. S1B), although
none of the bacteria were significant after multiple hypothesis
correction (Data Set S1, Sheet X). Although we did not identify a clear
signal to explain the delayed appearance of \emph{C. difficile} in the
5-day PEG mice that were allowed to recover for 10 days prior to
challenge, the delay is striking and could reflect changes in microbial
activity or metabolites that were not examined in this study.

\textbf{5-day laxative treatment does not promote more severe CDIs
despite altering the mucosal microbiota.} Given the findings from a
previous study that demonstrated that PEG treatment disrupts the mucus
layer and alters the immune response in mice (16), we decided to examine
the impact of PEG treatment on the mucosal microbiota and CDI severity.
To evaluate the mucosal microbiota, we sequenced snips of tissue
collected from the cecum, proximal colon, and distal colon. Similar to
what was observed with the stool samples, alpha diversity was lower in
the PEG-treated mice compared to clindamycin treatment (Fig. 3A).
Although alpha diversity continued to increase over time based on the
communities from PEG-treated mice collected at 20 and 30 days
post-infection (Fig. 3A). Group, day, and the interactions with other
variables (cage, experiment number, and sample type) explained the
majority of the variation observed in mucosal communities (PERMANOVA
combined R\textsuperscript{2} = 0.83, \emph{P} \textless{} 0.05, Fig.
3B, Data Set S1, Sheet X). \emph{Bacteroides}, \emph{Butyricoccus},
\emph{Clostridiales}, \emph{Clostridium Cluster XIVb},
\emph{Firmicutes}, and \emph{Ruminococcaceae} were consistently
different between treatment groups in both the tissue and stool
communities (Fig. 3C, Fig. 2E, Data Set S1, Sheet X). Next, we examined
CDI severity by evaluating cecum and colon histopathology (20) and found
there was no difference in cecum and colon scores between clindamycin
and PEG-treated mice that were challenged with \emph{C. difficile} at 4
days post-infection (Fig. 3D), the timepoint typically examined in
\emph{C. difficile} 630 challenged mice (ref). We also looked at 6 days
post-infection because that was when we started to see a large
difference in C\emph{. difficile} colonization levels between PEG- and
clindamycin-treated mice (Fig. 1C). Although, there was a slight
difference in the colon between PEG and clindamycin-treated mice, there
was no difference in the cecum and the overall score is still relatively
low given that the max possible summary score is 12 (Fig. 3E). Thus,
although PEG treatment had a profound impact on the mucosal microbiota,
the impact of \emph{C. difficile} on the cecum and colon was similar
between PEG and clindamycin treated mice.

\textbf{\emph{C. difficile} challenge does not have a synergistic
disruptive effect on the microbiota of PEG-treated mice} Because
\emph{C. difficile} itself can have an impact on the microbiota (21), we
also sequenced the tissue and stools of mock-challenged clindamycin and
5-day PEG treated mice. Examining the stools of the mock-infected PEG-
and clindamycin-treated mice revealed similar bacterial disruptions as
the \emph{C. difficile} challenged mice (Fig. S2A-C). Similarly, there
was no difference between the tissues of mock and \emph{C. difficile}
challenged mice (Fig. S2D-F). Thus, most of the microbiota alterations
we observed in the PEG-treated mice were a result of the laxative and
not an interaction between the laxative and \emph{C. difficile}.

\textbf{1-day laxative treatment results in transient \emph{C.
difficile} colonization and minor microbiota disruption}

\begin{itemize}
\tightlist
\item
  Draft of results

  \begin{itemize}
  \tightlist
  \item
    Explain motivation for experiment, set-up of experiment (Fig. 4A).
  \item
    Transient \emph{C. difficile} colonization in 1-day PEG treated mice
    with all PEG mice clearing by 7dpi (Fig. 4B)
  \item
    PCoA results (Day has larger R2 than treatment group). PEG mice are
    closer to baseline communites after 7-day period (Fig. 4C)
  \item
    Shannon results show alpha diversity is only transiently disrupted
    (Fig. 4D)
  \item
    Highlight bacteria that are disrupted by PEG or clindamycin
    treatment but recover within 7 days (Fig. 4E)
  \end{itemize}
\end{itemize}

\textbf{Post-CDI laxative treatment disrupts clearance in
clindamycin-treated mice regardless of whether an FMT is also
administered}

\begin{itemize}
\tightlist
\item
  Draft of results

  \begin{itemize}
  \tightlist
  \item
    Reiterate motivation for experiment. Saw only a transient microbiota
    disruption with 1-day PEG treatment alone. Also, see if adding FMT
    helps with clearance, mention groups have been using PEG or Abx/PEG
    treatments as a tool to engraft mice with human bacteria
    communities. Set up of experiment (Fig. 5A).
  \item
    Prolonged \emph{C. difficile} colonization in post-CDI PEG treated
    mice (Fig. 5B)
  \item
    FMT appears to partially restore alpha diversity (Shannon, but not
    richness Fig. 5C-D)
  \item
    PCoA (necessary?) Could comment on the clustering of Clindamycin and
    PEG-treated mice that received FMT (Fig. 5E)
  \item
    Only 2 genera significantly impacted by FMT treatment, likely not as
    important for clearance (Fig. 5F)
  \item
    Bacteria that consistently differ between groups over time,
    associated with prolonged colonization (Fig. G)
  \end{itemize}
\end{itemize}

\textbf{Five-day post-infection community data can predict mice that
will have prolonged \emph{C. difficile} colonization} After identifying
bacteria associated with the 5-day, 1-day and post-CDI 1-day PEG
treatments, we decided to examine the taxa that were influencing
prolonged \emph{C. difficile} colonization. We trained 3 types of
machine learning models (random forest, logistic regression, and support
vector machine) with input bacterial community data from 5 days
post-infection to predict whether the mice were colonized with \emph{C.
difficile} 10 days post-infection. We chose 5 days post-infection
because that was the earliest timepoint where we would see a treatment
effect in the mice that were given 1-day PEG treatment three days after
\emph{C. difficile} challenge and then administered an FMT or PBS
gavage. The random forest model had the highest performance (AUROC =
0.90, Data Set S1, Sheet X), so we next performed permutation importance
to examine the bacteria that were driving performance. We selected the
top 20 bacteria contributing to our models performance and examined
their relative abundance at 5 days post-infection, the timepoint used to
predict C. difficile colonization status on day 10 (Fig. 6A). Next, we
focused on the 5 genera that were \textgreater{} 1 \% relative abundance
in either the cleared or colonized mice and examined the relative
abundance dynamics of these bacteria over time. We found \emph{, }, and
\_ tended to have a higher relative abundance and \textbf{, } had a
lower relative abundance in the mice with prolonged colonization
compared to the mice that cleared \emph{C. difficile} (Fig. 6B).
Previous work examining the impact of PEG on the murine microbiota found
that PEG treatment resulted in the permanent loss of S24-7, also known
as \emph{Muribaculum intestinale} (16). We decided to check our
\emph{Porphyromonadaceae} OTUs because \emph{Muribaculum intestinale} is
known to be classified as \emph{Porphyromonadaceae} by the Ribosomal
Database Project (RDP) database (22), \emph{Porphyromonadaceae} was a
top feature in the random forest model predicting prolonged \emph{C.
difficile} colonization, and had a high relative abundance in the
communities of mice that cleared \emph{C. difficile} within 10 days. We
identified 4 OTUs that had at least (insert minimum percent identity) to
\emph{Muribaculum intestinale} and examined their abundance in mice that
either cleared or were still colonized with \emph{C. difficile} at 10
days post-infection (Fig. S3).

@@

\begin{itemize}
\item
  Summary of major findings (Fig. 7A)
\item
  Discussion of prolonged persistence. \emph{C. difficile} sequences
  detected in tissue samples. Association with mucin-degrading bacteria
  suggested by recent papers.
\item
  Discuss why we might not have observed more severe histology in PEG
  mice relative to clindamycin-treated mice

  \begin{itemize}
  \tightlist
  \item
    Antibiotics may also impact mucus layer
  \item
    Strain of bacteria used
  \end{itemize}
\item
  Protective bacteria missing in PEG-treated mice
\item
  Discuss what these findings might mean for human patients (Fig. 7B)

@@

\includegraphics{figure_5.pdf} A. Experimental timeline for groups
receiving PEG treatment post clindamycin gavage. B. Median and
individual CFU/g of C. difficile measured over time via culture.
Asterisks indicate time points with significant differences between
groups using Kruskall-Wallis tests with a Benjamini-Hochberg correction.
Limit of detection is 1e2 CFU/g. Background shading indicates PEG
treatment period for applicable groups. Opacity of points reflects time
point. C. Median and individual Shannon Diversity measured over time.
Asterisks, background shading, and opacity of points consistent with
(B). D. Median and individual richness measured over time. Asterisks,
background shading, and opacity of points consistent with (B).

\includegraphics{figure_6.pdf} \textbf{Figure 6. 1-day PEG treatment
post C. difficile challenge prolongs colonization regardless of whether
an FMT is also administered.} A. PCoA of all groups + FMT plotted over
time. Opacity of points reflects time point. B. Relative abundance of
genera significantly different over multiple time points post FMT or PBS
treatment, plotted over time. Limit of detection is .1\%. C. Relative
abundances of top 6 significant genera ranked by number of days
significant, plotted over time. Limit of detection is .1\%.

>>>>

\textbf{5-day laxative treatment leads to prolonged \emph{C. difficile}
colonization in mice.} We compared mice treated with the osmotic
laxative PEG 3350 to our standard 10 mg/kg clindamycin treatment, which
temporarily renders the mice susceptible to \emph{C. difficile}, with
mice typically clearing \emph{C. difficile} within 10 days
post-infection (9, 19). All PEG-treated mice were administered a 15\%
PEG solution in the drinking water for 5-days, one group was also
treated with clindamycin, and one group was allowed to recover for 10
days prior to challenge (Fig. 1A). PEG treatment resulted in weight loss
in all 3 groups of PEG-treated mice, with the greatest change in weight
observed on the fifth day of PEG treatment and the mice regained most of
the weight five days after treatment (Fig. 1B). After either PEG,
clindamycin, or PEG and clindamycin treatment all mice were challenged
with 10\textsuperscript{3} \emph{C. difficile} 630 spores (Fig. 1A). All
treatments rendered mice susceptible to \emph{C. difficile}
colonization. However, PEG-treated mice remained colonized with \emph{C.
difficile} at high levels through thirty days post-infection (Fig. 1C).
In contrast, the clindamycin-treated mice cleared \emph{C. difficile}
within ten days post-infection. Surprisingly, PEG-treated mice were
still susceptible to \emph{C. difficile} infection after 10-days of
recovery from treatment although \emph{C. difficile} was not detectable
in most of the group in the initial five days post-infection (Fig. 1C).
From 9 days post-infection onward, the median \emph{C. difficile}
stabilized for the 10-day recovery group of PEG-treated mice and
remained high through 30 days post-infection (Fig. 1C). Thus, osmotic
laxative treatment alone was sufficient to render mice susceptible to
prolonged \emph{C. difficile} colonization and PEG-treated mice remained
susceptible for up to ten days post-treatment.

\textbf{5-day laxative treatment differentially disrupts the fecal
microbiota compared to clindamycin treatment.} Since osmotic laxatives
and clindamycin have previously been shown to disrupt the murine
microbiota (ref), we hypothesized the different \emph{C. difficile}
colonization dynamics between mice treated with the osmotic laxative or
clindamycin were due to the two drugs having differential effects on the
microbiota. We profiled the stool microbiota over time by sequencing the
V4 region of the 16S rRNA gene to compare changes across treatment
groups. We found time and treatment group explained half of the observed
variation between fecal communities with most of the remaining variation
explained by interactions between treatment group and other experimental
variables including time, cage effects, and sequencing preparation
(PERMANOVA combined R\textsuperscript{2} = 0.95, \emph{P} \textless{}
0.001, Fig. 2A, Data Set S1, Sheet X). Cage effects refer to the
well-documented phenomenon that mice housed in the same cages have
similar microbial communities due to copraphagy (ref), we tried to
minimize the impact of cage effects on our experiment by breaking up
cagemates when assigning mice to treatment groups and primarily housing
only two mice per cage. Importantly, although we conducted a total of 5
separate experiments, the experiment number and it's interaction with
treatment group was not one of the variables that significantly
explained the observed variation in fecal communities (Data Set S1,
Sheet X). Interestingly, none of the treatment groups recovered to their
baseline community structure either 10 or 30 days post-infection
suggesting other community features besides recovery to baseline were
responsible for the prolonged \emph{C. difficile} colonization in
PEG-treated mice (Fig. 2B).

Next, we examined alpha diversity by looking at Shannon diversity index
over time. Although both clindamycin and PEG treatments decreased
diversity, Shannon diversity was lower in the groups of mice that
received PEG treatment compared to those that received clindamycin
through thirty days post-infection (Fig. 2C). We next examined the
bacterial genera that shifted after PEG treatment by comparing the
baseline samples of mice treated with only PEG to samples from the same
mice one day post-treatment. We found 18 genera that were altered by PEG
treatment (Data Set S1, Sheet X). The majority of the bacterial relative
abundances decreased after PEG, but \emph{Enterobacteriaceae} and
\emph{Bacteroides} increased and the increase in \emph{Bacteroides} was
unique to PEG treatment as \emph{Bacteroides} actually decreased in
clindamycin treated mice (Fig. 2D). Finally, we examined the bacteria
that differ across treatment groups over multiple timepoints. We found
24 genera were different over multiple timepoints out of the 33 that
were different between treatment groups (Fig. 2E, Data Set S1, Sheet X).
Thus, PEG has a significant impact on the fecal microbiota that was
maintained over time and distinct from clindamycin treatment.

Surprisingly, \emph{C. difficile} was not immediately detectable in the
stools of the PEG-treated mice that were allowed to recover for 10 days
prior to challenge. We decided to examine the bacteria that changed
during the post-infection period when the group median \emph{C.
difficile} shifted from undetectable at 1 day post-infection to
detectable in the stool samples with the median stabilizing around 8
days post-infection (Fig. S1A). Interestingly, we found
\emph{Erysipelotrichaceae}, \emph{Enterobacteriaceae}, and
\emph{Akkermansia} were changing during the time period when \emph{C.
difficile} was becoming detectable in the stools (Fig. S1B), although
none of the bacteria were significant after multiple hypothesis
correction (Data Set S1, Sheet X). Although we did not identify a clear
signal to explain the delayed appearance of \emph{C. difficile} in the
5-day PEG mice that were allowed to recover for 10 days prior to
challenge, the delay is striking and could reflect changes in microbial
activity or metabolites that were not examined in this study.

\textbf{5-day laxative treatment does not promote more severe CDIs
despite altering the mucosal microbiota.} Given the findings from a
previous study that demonstrated that PEG treatment disrupts the mucus
layer and alters the immune response in mice (16), we decided to examine
the impact of PEG treatment on the mucosal microbiota and CDI severity.
To evaluate the mucosal microbiota, we sequenced snips of tissue
collected from the cecum, proximal colon, and distal colon. Similar to
what was observed with the stool samples, alpha diversity was lower in
the PEG-treated mice compared to clindamycin treatment (Fig. 3A).
Although alpha diversity continued to increase over time based on the
communities from PEG-treated mice collected at 20 and 30 days
post-infection (Fig. 3A, Data Set S1, Sheet X). Group, day of the
experiment, and their interactions with other variables (cage,
experiment number, and sample type) explained the majority of the
variation observed in mucosal communities (PERMANOVA combined
R\textsuperscript{2} = 0.83, \emph{P} \textless{} 0.05, Fig. 3B, Data
Set S1, Sheet X). We saw the greatest difference in the mucosal
microbiota between treatment groups at 6 days post-infection with 16
(insert code instead) genera that were significantly different in all
three of the tissue types we collected (cecum, proximal colon, and
distal colon; Fig. S3A, Data Set S1, Sheet X). However, at 30 days
post-infection only \emph{Bacteroides}, \emph{Clostridiales},
\emph{Firmicutes}, and \emph{Ruminococcaceae} were different between
treatment groups and just in the cecum tissues (Fig. 3C, Fig. 2E, Data
Set S1, Sheet X). Interestingly, one of the genera that was different
between treatment groups at 6 days post-infection was
\emph{Peptostreptococcaceae} (the genera with sequences that match C.
difficile). \emph{Peptostreptococcaceae} was primarily only present in
the 5-day PEG treatment group of mice and decreased in the proximal and
distal colon tissues over time (Fig. S2B). Thus, PEG treatment had a
significant impact on the mucosal microbiota and we detected \emph{C.
difficile} sequences in the cecum, proximal colon, and distal colon
tissue communities.

Next, we examined CDI severity by evaluating cecum and colon
histopathology (20) and found there was no difference in cecum and colon
scores between clindamycin and PEG-treated mice that were challenged
with \emph{C. difficile} at 4 days post-infection (Fig. 3D), the
timepoint typically examined in \emph{C. difficile} 630 challenged mice
(ref). We also looked at 6 days post-infection because that was when we
started to see a large difference in C\emph{. difficile} colonization
levels between PEG- and clindamycin-treated mice (Fig. 1C). Although,
there was a slight difference in the colon between PEG and
clindamycin-treated mice, there was no difference in the cecum and the
overall score is still relatively low given that the max possible
summary score is 12 (Fig. 3E). Thus, although PEG treatment had a
profound impact on the mucosal microbiota, the impact of \emph{C.
difficile} 630 infection on the cecum and colon was similar between PEG
and clindamycin treated mice.

\textbf{\emph{C. difficile} challenge does not have a synergistic
disruptive effect on the microbiota of PEG-treated mice} Because
\emph{C. difficile} itself can have an impact on the microbiota (21), we
also sequenced the tissue and stools of mock-challenged clindamycin and
5-day PEG treated mice. Examining the stools of the mock-challenged mice
revealed similar bacterial disruptions as the \emph{C. difficile}
challenged mice (Fig. S2A-C). Similarly, there was no difference between
the tissues of mock and \emph{C. difficile} challenged mice (Fig.
S2D-F). Thus, most of the microbiota alterations we observed in the
PEG-treated mice were a result of the laxative and not an interaction
between the laxative and \emph{C. difficile}.

\textbf{1-day laxative treatment results in transient \emph{C.
difficile} colonization and minor microbiota disruption} Next, we
decided to examine how a shorter osmotic laxative perturbation would
impact the microbiome and susceptibility to \emph{C. difficile}. We
administered mice either a 1-day PEG treatment, a 1-day PEG treatment
with a 1-day recovery period, or a clindamycin treatment before
challenging them with \emph{C. difficile 630} (Fig. 3A). In contrast to
the 5-day PEG treated mice, the 1-day PEG treated mice were only
transiently colonized and cleared \emph{C. difficile} by 7 days
post-infection (Fig. 3B). The stool communities of PEG-treated mice were
also only transiently disrupted, with Shannon diversity close to
baseline by 7 days post-infection (Fig. 3C-D). We found 14 bacteria were
impacted by treatment but recovered close to baseline levels within 7
days post-infection including \emph{Enterobacteriaceae},
\emph{Clostridiales}, \emph{Porpyromonadaceae}, and
\emph{Ruminococcaceae} (Fig. 3E, Data Set S1, Sheet X). These findings
suggest the duration of treatment matters when considering the impact of
the laxatives, with shorter treatments resulting in a transient loss of
\emph{C. difficile} colonization resistance.

\textbf{Post-CDI laxative treatment disrupts clearance in
clindamycin-treated mice regardless of whether an FMT is also
administered} Since 1-day PEG treatment resulted in a more mild
microbiota perturbation, we decided to use the 1-day treatment to
examine the hypothesis that PEG helps to flush \emph{C. difficile}
spores from the intestine. We were also interested in exploring whether
PEG might help with engraftment in the context of FMTs. To examine the
impact of PEG treatment on \emph{C. difficile} clearance and FMT
treatment, we treated 4 groups of mice with clindamycin and then
challenged all mice with \emph{C. difficile} before administering the
following treatments: no additional treatment, 1-day PEG immediately
after challenge, and 1-day PEG treatment 3 days after challenge followed
by either administration of an FMT or PBS solution via gavage (Fig. 5A).
In contrast to our hypothesis, all groups of mice that received PEG
actually exhibited prolonged \emph{C. difficile} colonization (Fig. 5B).

\begin{itemize}
\item
  Draft of results

  \begin{itemize}
  \tightlist
  \item
    Prolonged \emph{C. difficile} colonization in post-CDI PEG treated
    mice (Fig. 5B)
  \item
    FMT appears to partially restore alpha diversity (Shannon, but not
    richness Fig. 5C-D)
  \item
    PCoA (necessary?) Could comment on the clustering of Clindamycin and
    PEG-treated mice that received FMT (Fig. 6A)
  \item
    Only 2 genera significantly impacted by FMT treatment, likely not as
    important for clearance (Fig. 6B)
  \item
    Bacteria that consistently differ between groups over time,
    associated with prolonged colonization (Fig. 6C)
  \end{itemize}
\end{itemize}

\textbf{Five-day post-infection community data can predict mice that
will have prolonged \emph{C. difficile} colonization} After identifying
bacteria associated with the 5-day, 1-day and post-CDI 1-day PEG
treatments, we decided to examine the taxa that were influencing
prolonged \emph{C. difficile} colonization. We trained 3 types of
machine learning models (random forest, logistic regression, and support
vector machine) with input bacterial community data from 5 days
post-infection to predict whether the mice were colonized with \emph{C.
difficile} 10 days post-infection. We chose 5 days post-infection
because that was the earliest timepoint where we would see a treatment
effect in the mice that were given 1-day PEG treatment three days after
\emph{C. difficile} challenge and then administered an FMT or PBS
gavage. The random forest model had the highest performance (AUROC =
0.90, Data Set S1, Sheet X), so we next performed permutation importance
to examine the bacteria that were driving performance. We selected the
top 10 bacteria contributing to our models performance (Fig. 7A) and
examined their relative abundance at 5 days post-infection, the
timepoint used to predict C. difficile colonization status on day 10
(Fig. 7B). Next, we focused on the 5 genera that were \textgreater{} 1
\% relative abundance in either the cleared or colonized mice and
examined the relative abundance dynamics of these bacteria over time. We
found \emph{, }, and \_ tended to have a higher relative abundance and
\textbf{, } had a lower relative abundance in the mice with prolonged
colonization compared to the mice that cleared \emph{C. difficile} (Fig.
7C). Previous work examining the impact of PEG on the murine microbiota
found that PEG treatment resulted in the permanent loss of S24-7, also
known as \emph{Muribaculum intestinale} (16). We decided to check our
\emph{Porphyromonadaceae} OTUs because \emph{Muribaculum intestinale} is
known to be classified as \emph{Porphyromonadaceae} by the Ribosomal
Database Project (RDP) database (22), \emph{Porphyromonadaceae} was a
top feature in the random forest model predicting prolonged \emph{C.
difficile} colonization, and had a high relative abundance in the
communities of mice that cleared \emph{C. difficile} within 10 days. We
identified 4 OTUs that had at least (insert minimum percent identity) to
\emph{Muribaculum intestinale} and examined their abundance in mice that
either cleared or were still colonized with \emph{C. difficile} at 10
days post-infection (Fig. S4).

@@

\begin{itemize}
\item
  Summary of major findings (Fig. 8A)
\item
  Discussion of prolonged persistence. \emph{C. difficile} sequences
  detected in tissue samples. Association with mucin-degrading bacteria
  suggested by recent papers.
\item
  Discuss why we might not have observed more severe histology in PEG
  mice relative to clindamycin-treated mice

  \begin{itemize}
  \tightlist
  \item
    Antibiotics may also impact mucus layer
  \item
    Strain of bacteria used
  \end{itemize}
\item
  Protective bacteria missing in PEG-treated mice
\item
  Discuss what these findings might mean for human patients (Fig. 8B)

@@

\includegraphics{figure_5.pdf} \textbf{Figure 5. 1-day PEG treatment
post \emph{C. difficile} challenge prolongs colonization regardless of
whether an FMT is also administered.} All mice were administered 10
mg/kg clindamycin intraperitoneally (IP) 1 day before challenge with
\emph{C. difficile} 630 spores on day 0. 15\% PEG 3350 was administered
in the drinking water for 1 day. Mice feces were collected daily through
the end of the experiment (30 days post-infection for Clind. + 1-day PEG
group, 15 days post-infection for all others). CFU quantification data
was not available for each mouse due to stool sampling difficulties
(particularly the day the mice came off of the PEG treatment) or early
deaths. B. CFU/g of \emph{C. difficile} stool measured over time via
serial dilutions. C. Median and individual Shannon Diversity measured
over time. D. Median and individual richness measured over time. B-D.
Asterisks indicate time points with significant differences between
groups using Kruskall-Wallis tests with a Benjamini-Hochberg correction
for testing multiple times points. Black line represents the limit of
detection for the first serial dilution. Background shading indicates
PEG treatment period for applicable groups. Opacity of points reflects
time point. The data presented are from a total of *** separate
experiments. \newpage

\includegraphics{figure_6.pdf} \textbf{Figure 6. For 1-day PEG treatment
post \emph{C. difficile} challenge mice that also receive an FMT only
some bacterial genera were restored.} A. PCoA of all groups + FMT
plotted over time. Opacity of points reflects time point. B. Relative
abundance of genera significantly different over multiple time points
post FMT or PBS treatment, plotted over time. Limit of detection is
.1\%. C. Relative abundances of top 6 significant genera ranked by
number of days significant, plotted over time. Limit of detection is
.1\%.
